Inaction on climate change could cost one year of life in some European countries

### 2018-10-23 MS-cclifeexpec.Rmd
minor edits.

---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: xelatex
    template: LATEX/svm-latex-ms.tex
title: "Inaction on climate change could approach one year of life in some European countries"
thanks: "All data and code that supports these conclusions are available as supplementary materials."
author:
- name: Mathew E. Hauer^1^*
  affiliation: University of Georgia
- name: Alexis R. Santos^2^
  affiliation: Pennsylvania State University

abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change. However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, and demographic groups. No studies have comprehensively connected excess mortality due to climate change to anticipated reductions in life expectancy. Without properly situating the potential loss of life within the contexts of life expectancy, we risk misrepresenting cost of climate change on mortality. In this paper, we convert excess mortality estimates due to increaes in extreme weather from climate change into potential reductions of life expectancy at birth and report the cost of climate change on human longevity. We find climate change extremes could become the third largest life expectancy reducer behind heart disease and cancer in some European countries. Thus, the cost of inaction on climate change could approach one year of life in some countries and possibly exceed that."

keywords: "Climate change, Life expectancy, Mortality, Demography, Excess mortality, Europe, Public health"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 12pt
spacing: double
bibliography: LATEX/ccmort.bib
biblio-style: apsr
# use apsr or nature
fig_caption: true
header-includes:
- \usepackage[all]{nowidow}
- \usepackage{rotating}
- \usepackage{fancyhdr}
- \usepackage[table]{xcolor}
- \usepackage{tabularx}
- \usepackage{makecell}
- \usepackage{xcolor}
- \pagestyle{fancy}
- \fancyhead{}
- \fancyhead[C]{Climate change life expectancy}

@@

* Corresponding author. hauer@uga.edu. p: 706-542-9369.

^1^ Department of Sociology, Florida State University. 201 N. Milledge Ave. Athens, GA USA 30602.

^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802.

@@

Climate change's implications on humanity go far beyond estimates of economic damages [@hsiang2017estimating], estimates of displacement [@rigaud2018groundswell], or human conflict [@barnett2007climate]  but have the potential to lead to the loss of human life [@forzieri2017increasing;@pachauri2014climate]. As impact quantification studies move further from the physical sciences of climate change into the social sciences, properly quantifying and conveying the impact of climate change on public health is of increasing importance [@melillo2014climate;@cloyd2016engagement].

Scholars have long estimated the mortality risks associated with climate change and typically use excess or extra mortality [@forzieri2017increasing;@wilson2017climate;@mcmichael2006climate;@zanobetti2012summer].

Although such estimates are useful, the excess mortality estimates are rather sterile -- one death is a tragedy, a million deaths a statistic -- and difficult to relate to on an individual level. Life expectancy at birth ($e_0$), on the other hand, provides potent comparisons of mortality vectors and converts excess mortality into an intuitively understood metric, relatable to everyone. To our knowledge, no studies exist that comprehensively examine the potential reductions of life expectancy due to climate change. Without properly situating the potential loss of life within the contexts of metrics such as life expectancy, we risk underestimating the impact of climate change on human mortality.

Using published data on excess mortality [@forzieri2017increasing], we connect climate change excess mortality to life expectancy in a mortality model. By estimating the increase in age-specific mortality rates associated with previously published excess death estimates, we can assess how much climate change could reduce the anticipated longevity of the average person in twenty-eight European countries. This approach allows us to quantify the impact of climate change on human longevity and answer the following questions: What is the cost of inaction on climate change on human longevity? and How does climate change related mortality compare to other mortality vectors? Our results situate climate change mortality within the broader context of human mortality and can be used to inform public health interventions to prevent such futures.

Our results suggest that climate change could emerge as a potent new mortality vector in the 20th century. Mitigation strategies (ie reductions in greenhouse gas emisions) and adaptation strategies (ie outreach efforts, retrofitting public buildings, etc.) would help prevent this new public health concern.

# Background
## Climate change and human mortality

# Methods and Materials
We estimate changes in life expectancy by combining three primary datasets: excess mortality data due to meterological hazards related to climate change from Forzieri et al. [@forzieri2017increasing], cause-of-death distributions from the Global Burden of Disease study (GBD) [@GBD;@wang2012age], and life tables from the Human Mortality Database (HMD) [@HMD]. Forzieri et al. [@forzieri2017increasing] focused on the projected excess mortality of heatwaves, coldwaves, wildfires, droughts, river and coastal floods, and windstorms under the business-as-usual greenhouse gas emission scenario (SRES A1B). They modelled long-term, small-area demographic changes in European countries that correspond to a middle-of-the-road socioeconomic scenario (SSP2 with medium fertility, medium mortality, medium migration, and the Global Education Trend education scenario) and based their projected mortality on extrapolations of an exhaustive examination of contemporary disaster databases within spatially rectified, high-resolution gridded datasets. While the disaster data are not standardized across country, Forzeri et al. imputed data in incomplete time periods and countries. This imputation could mask spatial variability at the sub-national level, but here we use their country-aggregated results. Their results represent business-as-usual climate change and human development without incorporating potential adaptation. They found a potential 150,000+ climate change related fatalities per year by the mid 2080s with climate change contributing to 90% of mortality rise as opposed to population changes. These data provide the anticipated fatalities but not the effect these deaths might have on life expectancy.

To convert excess mortality to life expectancy, we use data gathered from the HMD [@HMD] for corresponding European countries in five-year life tables in conjunction with cause-specific mortality data from the GBD [@GBD;@wang2012age]. The HMD comprises only complete, official vital event statistics and represents the most accurate and complete compilation of mortality data available. We allocate excess multi-hazard mortality for each country from supplementary table S8 in Forzieri et al. [@forzieri2017increasing] based on the observed mortality schedule for environmental heat and cold exposure deaths from the GBD for each country in the previous decade.

\begin{table}
\caption{\textbf{Most recent mortality data available by country in the HMD.}} \label{Table2}
\begin{tabular}{ll|ll|ll|ll}
Country & Year & Country & Year & Country & Year & Country & Year \\
\hline
AUT     & 2014 & ESP     & 2014 & HUN     & 2014 & NLD     & 2016 \\
BEL     & 2015 & EST     & 2014 & IRL     & 2014 & NOR     & 2014 \\
BGR     & 2010 & FIN     & 2015 & ISL     & 2016 & POL     & 2014 \\
CHE     & 2016 & FRATNP  & 2015 & ITA     & 2014 & PRT     & 2015 \\
CZE     & 2016 & GBR     & 2016 & LTU     & 2014 & SVK     & 2014 \\
DEUT    & 2015 & GRC     & 2013 & LUX     & 2014 & SVN     & 2014 \\
DNK     & 2016 & HRV     & 2016 & LVA     & 2014 & SWE     & 2016 \\
\hline
\end{tabular}
\end{table}

We then recalculated age-specific mortality rates in five-year life tables by adding the climate change mortality onto the underlying mortality data and create four subsequent life tables ($baseline$, $low$, $mid$, $high$) using standard life table techniques [@wunsch2013life] based on the published confidence intervals from [@forzieri2017increasing]. We used a variant of the cause-deleted life tables approach [@brand2005approximations;@beltran2008integrated] to measure the impact of excess mortality on life expectancy at birth or $e_0$. Changes in life expectancy are reported based on differentials from $e_0$ between the 2080s and the most recent complete life table in the HMD (see **Table 1** for the years used). Complete life tables for all countries are available in the *Supplementary Dataset* and all computer code to replicate our results are available online[^1].

[^1]: Replication files are located here: <https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1>

## Data
To create life expectancy differentials, we used three sources of data. The first is the baseline life table data. Baseline life tables comes from the Human Mortality Database (HMD) [@HMD]. These datasets comprise only official vital event registration and are considered the most complete and accurate source of mortality data available. We downloaded the most recent 5x1 (by age and year) life table data for twenty-eight European countries. We used the following data elements from the five-year age group life table data in the HMD:

The second dataset comes from published projections of climate-related mortality in Europe [@forzieri2017increasing]. Forzieri et al. [@forzieri2017increasing] combined data from the International Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool (NatCatSERVICE) to create exposure and fatality statistics related to six weather-related hazards -- heatwaves, cold waves, droughts, wildfires, river and coastal floods, and windstorms. They downscaled these exposures to  1km grid scales and integrated them into small-area demographic projections for Europe out to 2100. Heatwaves and coldwaves account for the vast majority of projected fatalities. The number of future anticipated deaths are available in their Supplementary Table S8 [@forzieri2017increasing] and provide the magnitude of deaths for our study.

The third dataset comes from the GBD [@GBD;@wang2012age]. The GBD is the most comprehensive worldwide dataset of epidemiological data produced and provides cause-specific mortality by age/sex/geography/year for 249 causes of death in 195 countries and territories. We gathered data on age-specific deaths and mortality rates ($m_x$) for mortality from environmental heat and cold exposure (cause C.2.9) for the period 2006-2015 for each country in the study. These data provide a mortality schedule to fit the projected climate change excess mortality from Fozieri et al [@forzieri2017increasing]. Thus, we assume that the age-specific mortality schedules observed between 2006 and 2015 due to environmental heat and cold exposure in the GBD (see \autoref{suppfig}) are likely to remain unchanged in the future.

@@

## Methods
We derived a number of additional variables to accomplish our analysis. First, we abridged the life table data from the HMD from ages 100+ to 80+ to conform to the GBD cause-specific mortality schedules.

@@

To determine the differences between life expectancy compared to the baseline, we simply subtract $e_{x,i,s}$ (for scenario $LOW$, $MID$, and $HIGH$) from $e_{x,i,BASE}$ as described by [@beltran2008integrated].

# Results
We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` to `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**). This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States for the year 2000. Although the average European country could see a change of `r round(mean(lt_lancet$DIF_MID),2)` years, several countries are likely to experience considerably greater reductions in life expectancy. Luxembourg could see a reduction of up to `r round(lt_lancet$DIF_HIGH[which(lt_lancet$CNTRY == "LUX")],2)*-1` years and the medium-variant of climate hazards could cost the average Spaniard `r round(lt_lancet$DIF_MID[which(lt_lancet$CNTRY == "ESP")],2)*-1` years of life.

@@

**\autoref{Table1}** reports the rankings of Age Standardised Death Rates (ASDR) for the leading causes of death in these European countries (ASDR's for other CODs come from Eurostat (online data code: hlth_cd_asdr2)) with our estimates of climate change included. In many European countries, climate change could emerge as a top five killer. Only combined CODs, such as all circulatory diseases or all cancers, exhibit higher ASDRs if climatic hazards are more severe than anticipated for Spain, Luxembourg, and France.

@@

# Discussion
In this article, we demonstrate the impact climate change could have on life expectancy at birth in twenty-eight European countries. Previous studies on climate change and excess mortality potentially miscommunicate the impact climate change could have on human mortality. Contrary to excess mortality estimates, life expectancy is routinely used as a primary metric for communicating overall health outcomes and enjoys widespread use by major international organizations [@world2015world;@marmot2012building;@salomon2012healthy]. Life expectancy and its derivatives are the recommended metrics for population health [@parrish2010peer]. Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into individual outcomes. Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].

@@

\begin{sidewaystable}

\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} \label{Table1}
\rowcolors{2}{gray!25}{white}
\small
\begin{tabularx}{\textwidth}{l|XXXXXXXXX}
  \hline
   \rowcolor{white}
 Country & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\ 
  \hline
Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 78.05\\\cellcolor{blue!25}\begin{tiny}(63.26 - 101.6)\end{tiny}} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis. of the\\Nervous Sys\\ 48.5} & \makecell{Lung\\Cancer\\ 47.8} & \makecell{Colorectal\\Cancer\\ 33.6} & \makecell{Suicide\\ 8.2} & \makecell{Transport\\Accidents\\ 4.3} \\ 
Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 76.93\\\cellcolor{blue!25}\begin{tiny}(2.36 - 160.9)\end{tiny}} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis. of the\\Nervous Sys\\ 38} & \makecell{Colorectal\\Cancer\\ 25.5} & \makecell{Suicide\\ 13.4} & \makecell{Transport\\Accidents\\ 6} \\ 
Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.87\\\cellcolor{blue!25}\begin{tiny}(28.15 - 68.2)\end{tiny}} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis. of the\\Nervous Sys\\ 34.3} & \makecell{Colorectal\\Cancer\\ 27} & \makecell{Suicide\\ 6.3} & \makecell{Transport\\Accidents\\ 5.6} \\ 
France & \makecell{Respiratory\\Diseases\\ 52} & \makecell{Dis. of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 49.61\\\cellcolor{blue!25}\begin{tiny}(18.24 - 93.5)\end{tiny}} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\ 
Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 35.95\\\cellcolor{blue!25}\begin{tiny}(29.93 - 49.4)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis. of the\\Nervous Sys\\ 32.8} & \makecell{Suicide\\ 11.3} & \makecell{Transport\\Accidents\\ 7.8} \\ 
Ireland & \makecell{Heart\\Disease\\ 147.5} & \makecell{Respiratory\\Diseases\\ 125.9} & \makecell{Lung\\Cancer\\ 61.5} & \makecell{Dis. of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 44.65\\\cellcolor{blue!25}\begin{tiny}(13.57 - 119.6)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\ 
Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}Climate\\\cellcolor{blue!25}Change\\\cellcolor{blue!25} 25.02\\\cellcolor{blue!25}\begin{tiny}(11.19 - 37.8)\end{tiny}} & \makecell{Dis. of the\\Nervous Sys\\ 21.1} & \makecell{Suicide\\ 18.9} & \makecell{Transport\\Accidents\\ 6.7} \\ 

>>>>

---
output: 
  pdf_document:
    # citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: LATEX/svm-latex-ms.tex
title: "Inaction on climate change could cost one year of life in some European countries"
thanks: "All data and code that supports these conclusions are available as supplementary materials. The authors would like to thank E. Carlson for his helpful comments."
author:
- name: Mathew E. Hauer^1,2^*
  affiliation: Florida State University
- name: Alexis R. Santos^3^
  affiliation: Pennsylvania State University

abstract: "Climate change-related excess mortality estimates clearly demonstrate a dramatic impact on public health and human mortality. However, life expectancy at birth is more easily communicated and understood by the public. By properly situating climate change mortality within the contexts of life expectancy, we better represent the cost of climate change on longevity. In this paper, we convert excess mortality estimates due to increases in extreme weather from climate change into potential reductions in life expectancy at birth in twenty-eight European countries. We project climate change extremes to reduce life expectancy at birth by 0.24 years for the average European country with differences in excess of 1.0 years in some countries by 2100. We only estimate the impact of mortality directly related to climate extremes, making our estimates conservative. Thus, the cost of inaction on climate change could approach, and likely to exceed, one year of life in some European countries."

# "Climate change-related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change. However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, age, and demographic groups. Without properly situating the potential loss of life within the contexts of life expectancy, we risk misrepresenting the cost of climate change on mortality. In this paper, we convert excess mortality estimates due to increases in extreme weather from climate change into potential reductions of life expectancy at birth in twenty-eight European countries. We project climate change extremes to reduce life expectancy at birth by 0.24 years for the average European country but could exceed 1.0 years in some by 2100. These results are conservative when compared to the broader impact of climate change on human longevity as we only estimate the impact of mortality directly related to climate extremes. Thus, the cost of inaction on climate change could approach one year of life in some countries and likely to exceed that."
keywords: "Climate change, Life expectancy, Mortality, Demography, Excess mortality, Europe, Public health"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 12pt
spacing: double
bibliography: LATEX/ccmort.bib
# biblio-style: apsr
csl : LATEX/american-sociological-association.csl
# use apsr or nature
fig_caption: true
header-includes:
- \usepackage[all]{nowidow}
- \usepackage{rotating}
- \usepackage{fancyhdr}
- \usepackage[table]{xcolor}
- \usepackage{tabularx}
- \usepackage{makecell}
- \usepackage{xcolor}
- \pagestyle{fancy}
- \fancyhead{}
- \fancyhead[C]{Climate change life expectancy}

@@

* Corresponding author. mehauer@fsu.edu p: 850-644-7103.

^1^ Department of Sociology, Florida State University. 113 Colleggiate Way, Bellamy Building, Tallahassee FL 30602.

^2^ Center for Demography and Population Health, Florida State University.

^3^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802.

@@

Climate change's implications on humanity go far beyond estimates of economic damages [@hsiang2017estimating], estimates of displacement [@rigaud2018groundswell], or human conflict [@barnett2007climate]  but have the potential to contribute to the loss of human life [@forzieri2017increasing;@pachauri2014climate]. As impact quantification studies move further from the physical sciences of climate change, properly quantifying and conveying the impact of climate change on public health is of increasing importance [@melillo2014climate;@cloyd2016engagement].

Scholars have long estimated the mortality risks associated with climate change and typically use excess or extra mortality [@forzieri2017increasing;@wilson2017climate;@mcmichael2006climate;@zanobetti2012summer]. For example, Forzieri et al. [-@forzieri2017increasing] projected climate change extremes to increase European mortality by 150,000+ persons per year by the end of the century. Although such estimates are useful, the excess mortality estimates are rather sterile -- one death is a tragedy, a million deaths a statistic -- and difficult to relate to on a personal level. Life expectancy at birth ($e_0$), on the other hand, provides potent comparisons of mortality vectors and converts excess mortality into an intuitively understood metric, relatable to everyone. To our knowledge, no studies exist that comprehensively examine the potential reductions of life expectancy due to climate change. Without properly situating the potential loss of life within the contexts of metrics such as life expectancy, we risk underestimating the impact of climate change on human mortality.

Using published data on excess mortality [@forzieri2017increasing], we connect climate change excess mortality to life expectancy in a mortality model. By estimating the increase in age-specific mortality rates associated with previously published excess death estimates, we can assess how much climate change could reduce the anticipated longevity of the average person in twenty-eight European countries. This approach allows us to quantify the impact of climate change on human longevity and answer the following question: What is the cost of inaction on climate change on human longevity? Our results situate climate change mortality within the broader context of human mortality and can be used to inform public health interventions to prevent such futures.

Our results suggest that climate change could emerge as a potent public health threat in the 20th century. Mitigation strategies (ie reductions in greenhouse gas emisions) and adaptation strategies (ie outreach efforts, retrofitting public buildings, etc.) would help prevent this new public health concern.

# Methods and Materials

## Data

We estimate changes in life expectancy by combining three primary datasets: excess mortality data due to meterological hazards related to climate change from Forzieri et al. [-@forzieri2017increasing], cause-of-death distributions from the Global Burden of Disease study (GBD) [@GBD;@wang2012age], and life tables from the Human Mortality Database (HMD) [@HMD].

Forzieri et al. [-@forzieri2017increasing] combined data from the International Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool (NatCatSERVICE) to create exposure and fatality statistics related to six weather-related hazards -- heatwaves, cold waves, droughts, wildfires, river and coastal floods, and windstorms -- and estimated excess mortality associated with those hazards during a baseline period of 1981-2010 in 28 European countries. They only considered fatalities *directly attributable* to the event and *exclude* increased deaths from common causes that were observed to rise during the event, such as cardiovascular or respiratory deaths[^1]. While the disaster data are not standardized across country, Forzeri et al. imputed data in incomplete time periods and countries. This imputation could mask spatial variability at the sub-national level, but here we use their country-aggregated results. They downscaled these exposures rates to 1km grid cells and integrated them into small-area demographic projections for Europe out to 2100 that correspond to a middle-of-the-road socioeconomic scenario (SSP2 with medium fertility, medium mortality, medium migration, and the Global Education Trend education scenario [@jiang2014internal;@samir2017human;@o2017roads]) and based their projected mortality on extrapolations of their calculated exposures. Their results represent business-as-usual climate change and human development without incorporating potential adaptations to climate change. They found a potential 150,000+ climate change related fatalities per year by the mid 2080s with climate change contributing to 90% of mortality rise as opposed to population changes. The number of future anticipated deaths are available in their Supplementary Table S8 [@forzieri2017increasing] and provide the magnitude of deaths for our study.

[^1]: "The fatality data from the two databases considered is likely to not include increased deaths from common causes that were observed to rise at the population level but for which individual deaths could not be attributed to the event. For example,... increased risk of cardiovascular and respiratory deaths ... may be severly underreported in the EMDAT and NatCatSERVICE." [see @forzieri2017increasing, pp. 9, *Supplementary Materials*].

@@

The Forzieri et al. [-@forzieri2017increasing] data contain only mortality totals and report no age detail, precluding a direct conversion from excess mortality to life expectancy. To convert excess mortality to life expectancy, we allocate excess multi-hazard mortality for each country based on the observed age-specific mortality schedule for environmental heat and cold exposure deaths from the GBD [@GBD;@wang2012age]. Forzieri et al. found that extreme heat (as opposed to other five weather-related hazards) account for 99% of the anticipated excess mortality by the end of the century. The GBD is the most comprehensive worldwide dataset of epidemiological data produced and provides cause-specific mortality by age/sex/geography/year for 249 causes of death in 195 countries and territories. We gathered data on age-specific deaths and mortality rates ($m_x$) for mortality from environmental heat and cold exposure (cause C.2.9) for the period 2006-2015 for each country in the study. These data provide a mortality schedule to fit the projected climate change excess mortality from Fozieri et al [@forzieri2017increasing]. Thus, we assume that the age-specific mortality schedules observed between 2006 and 2015 due to environmental heat and cold exposure in the GBD (see \autoref{suppfig}) are likely to remain unchanged in the future. We only use the GBD to distribute the Forzieri et al. deaths by age.

With the Forzier et al data and the GBD we estimate the anticipated age-specific mortality schedules due to climate change in twenty-eight European countries but we cannot compare life expectancy differentials in the absence of Climate Change using those two datasets alone. To create baseline mortality schedules and life expectancy, we gather data from the Human Mortality Database [@HMD] for the corresponding European countries. The HMD comprises only complete, official vital event statistics and represents the most accurate and complete compilation of mortality data available. We downloaded the most recent 5x1 (age by year) life table data for the twenty-eight European countries (see \autoref{Table2} for a list of countries and most recent mortality data).

\begin{table}
\caption{\textbf{Most recent mortality data available by country in the HMD.}} \label{Table2}
\begin{tabular}{ll|ll|ll|ll}
Country & Year & Country & Year & Country & Year & Country & Year \\
\hline
AUT     & 2014 & ESP     & 2014 & HUN     & 2014 & NLD     & 2016 \\
BEL     & 2015 & EST     & 2014 & IRL     & 2014 & NOR     & 2014 \\
BGR     & 2010 & FIN     & 2015 & ISL     & 2016 & POL     & 2014 \\
CHE     & 2016 & FRATNP  & 2015 & ITA     & 2014 & PRT     & 2015 \\
CZE     & 2016 & GBR     & 2016 & LTU     & 2014 & SVK     & 2014 \\
DEUT    & 2015 & GRC     & 2013 & LUX     & 2014 & SVN     & 2014 \\
DNK     & 2016 & HRV     & 2016 & LVA     & 2014 & SWE     & 2016 \\
\hline
\end{tabular}
\end{table}
We used the following data elements from the five-year age group life table data in the HMD:

## Methods

We derived a number of additional variables to accomplish our analysis. First, we abridged the life table data from the HMD from ages 100+ to 80+ to conform to the GBD cause-specific mortality schedules.

@@

To determine the differences between life expectancy compared to the baseline, we simply subtract $e_{x,i,s}$ (for scenario $LOW$, $MID$, and $HIGH$) from $e_{x,i,BASE}$ as described by [@beltran2008integrated].Complete life tables for all countries are available in the *Supplementary Dataset* and all computer code to replicate our results are available online[^2].

[^2]: Replication files are located here: <https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1>

# Results
We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` to `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**) by 2100. This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States for the year 2000. Although the average European country could see a change of `r round(mean(lt_lancet$DIF_MID),2)` years, several countries are likely to experience considerably greater reductions in life expectancy. Luxembourg could see a reduction of up to `r round(lt_lancet$DIF_HIGH[which(lt_lancet$CNTRY == "LUX")],2)*-1` years and the medium-variant of climate hazards could cost the average Spaniard `r round(lt_lancet$DIF_MID[which(lt_lancet$CNTRY == "ESP")],2)*-1` years of life.

@@

<!-- **\autoref{Table1}** reports the rankings of Age Standardised Death Rates (ASDR) for the leading causes of death in these European countries (ASDR's for other CODs come from Eurostat (online data code: hlth_cd_asdr2)) with our estimates of climate change included. In many European countries, climate change could emerge as a top five killer. Only combined CODs, such as all circulatory diseases or all cancers, exhibit higher ASDRs if climatic hazards are more severe than anticipated for Spain, Luxembourg, and France. -->

@@

# Discussion
In this article, we demonstrate the impact climate change could have on life expectancy at birth due to extreme weather events in twenty-eight European countries. Previous studies on climate change and excess mortality potentially miscommunicate the impact climate change could have on human mortality. Contrary to excess mortality estimates, life expectancy is routinely used as a primary metric for communicating overall health outcomes and enjoys widespread use by major international organizations [@world2015world;@marmot2012building;@salomon2012healthy]. Life expectancy and its derivatives are the recommended metrics for population health [@parrish2010peer]. Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into personably relatable outcomes. Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].

@@

<!-- \begin{sidewaystable} -->

<!-- \caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} \label{Table1} -->
<!-- \rowcolors{2}{gray!25}{white} -->
<!-- \small -->
<!-- \begin{tabularx}{\textwidth}{l|XXXXXXXXX} -->
<!--   \hline -->
<!--    \rowcolor{white} -->
<!--  Country & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\  -->
<!--   \hline -->
<!-- Spain & \makecell{Respiratory\\Diseases\\ 91.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 78.05\\\cellcolor{blue!25}\begin{tiny}(63.26 - 101.6)\end{tiny}} & \makecell{Heart\\Disease\\ 68.2} & \makecell{Dis. of the\\Nervous Sys\\ 48.5} & \makecell{Lung\\Cancer\\ 47.8} & \makecell{Colorectal\\Cancer\\ 33.6} & \makecell{Suicide\\ 8.2} & \makecell{Transport\\Accidents\\ 4.3} \\  -->
<!-- Luxembourg & \makecell{Heart\\Disease\\ 80.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 76.93\\\cellcolor{blue!25}\begin{tiny}(2.36 - 160.9)\end{tiny}} & \makecell{Respiratory\\Diseases\\ 63.8} & \makecell{Lung\\Cancer\\ 59.6} & \makecell{Dis. of the\\Nervous Sys\\ 38} & \makecell{Colorectal\\Cancer\\ 25.5} & \makecell{Suicide\\ 13.4} & \makecell{Transport\\Accidents\\ 6} \\  -->
<!-- Italy & \makecell{Heart\\Disease\\ 98.3} & \makecell{Respiratory\\Diseases\\ 58.3} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 49.87\\\cellcolor{blue!25}\begin{tiny}(28.15 - 68.2)\end{tiny}} & \makecell{Lung\\Cancer\\ 49.4} & \makecell{Dis. of the\\Nervous Sys\\ 34.3} & \makecell{Colorectal\\Cancer\\ 27} & \makecell{Suicide\\ 6.3} & \makecell{Transport\\Accidents\\ 5.6} \\  -->
<!-- France & \makecell{Respiratory\\Diseases\\ 52} & \makecell{Dis. of the\\Nervous Sys\\ 50.2} & \makecell{Lung\\Cancer\\ 50.1} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 49.61\\\cellcolor{blue!25}\begin{tiny}(18.24 - 93.5)\end{tiny}} & \makecell{Heart\\Disease\\ 49.3} & \makecell{Colorectal\\Cancer\\ 26.1} & \makecell{Suicide\\ 14.1} & \makecell{Transport\\Accidents\\ 5.1} \\  -->
<!-- Portugal & \makecell{Respiratory\\Diseases\\ 116.7} & \makecell{Heart\\Disease\\ 69.6} & \makecell{Lung\\Cancer\\ 36.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 35.95\\\cellcolor{blue!25}\begin{tiny}(29.93 - 49.4)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 35} & \makecell{Dis. of the\\Nervous Sys\\ 32.8} & \makecell{Suicide\\ 11.3} & \makecell{Transport\\Accidents\\ 7.8} \\  -->
<!-- Ireland & \makecell{Heart\\Disease\\ 147.5} & \makecell{Respiratory\\Diseases\\ 125.9} & \makecell{Lung\\Cancer\\ 61.5} & \makecell{Dis. of the\\Nervous Sys\\ 48.7} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 44.65\\\cellcolor{blue!25}\begin{tiny}(13.57 - 119.6)\end{tiny}} & \makecell{Colorectal\\Cancer\\ 32.4} & \makecell{Suicide\\ 11} & \makecell{Transport\\Accidents\\ 4} \\  -->
<!-- Slovenia & \makecell{Heart\\Disease\\ 102.8} & \makecell{Respiratory\\Diseases\\ 66.3} & \makecell{Lung\\Cancer\\ 58.6} & \makecell{Colorectal\\Cancer\\ 38.4} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 25.02\\\cellcolor{blue!25}\begin{tiny}(11.19 - 37.8)\end{tiny}} & \makecell{Dis. of the\\Nervous Sys\\ 21.1} & \makecell{Suicide\\ 18.9} & \makecell{Transport\\Accidents\\ 6.7} \\  -->
<!-- Croatia & \makecell{Heart\\Disease\\ 306.5} & \makecell{Lung\\Cancer\\ 65.2} & \makecell{Respiratory\\Diseases\\ 59.7} & \makecell{Colorectal\\Cancer\\ 51} & \cellcolor{blue!25}\makecell{\cellcolor{blue!25}CC\\\cellcolor{blue!25}Events\\\cellcolor{blue!25} 24.44\\\cellcolor{blue!25}\begin{tiny}(13.32 - 38.6)\end{tiny}} & \makecell{Dis. of the\\Nervous Sys\\ 21.3} & \makecell{Suicide\\ 16.8} & \makecell{Transport\\Accidents\\ 8.9} \\  -->
